---
title: "Phylogenetic Least Squares analysis"
author: "O. Johnson"
date: "2024-09-29"
output: html_document
---

# About this script

The following document provides the R code for the analysis for the paper: Johnson *et al.* 

**\*Corresponding author:**
[olivia.johsnon\@utas.edu.au](mailto:olivia.johsnon@utas.edu.au){.email}

+------------------------+-------------------------------------------------------------------------------------------------------+
| **R-version**          | 4.4.1 (2024-06-14 ucrt) -- "Race for Your Life"                                                        |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **platform**           | x86_64-w64-mingw32/x64                                                                                   |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Article DOI**        |                                                                                                       |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Article link**       |                                                                                                       |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Article citation**   | Johnson *et al.*  |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Time series**        | 1992 - 2024                                                                                          |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Geographical scale** | Continental Australia                                                                                 |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Code author contact**| olivia.johnson@utas.edu.au                                                                            |
+------------------------+-------------------------------------------------------------------------------------------------------+

# Set-up

```{r knitr-options}
#| include = FALSE

library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

# save csv files for each step?
save_csv <- TRUE

```

# Loading packages
Load in all required packages for this analysis
```{r packages}
library(knitr)
library(tidyverse)
library(janitor)
library(lubridate)
library(dplyr)
library(data.table)
library(caper)
library(ape)
library(fishtree)
library(readr)
library(rotl)
library(taxize)
library(phytools)
library(ggtree)
library(httr)
library(jsonlite)
library(ape)           
library(ggtree)        
library(dplyr)         
library(ggplot2)

# "not in" function
`%!in%` <- Negate(`%in%`)
```

#Read in data
```{r - data input}

species <- read_csv("LMM_output.csv")

covariates <- read_csv("covariates_final.csv") %>%
  rename(site_code = "Name")

lats_and_longs <- read_csv("all_lats_longs_clean.csv")

```

```{r - Phyolgeny of species}

###################################
# Fetch OTT IDs for species

# Extract species names and match with Open Tree of Life
resolved_names <- species %>%
  pull(species_name) %>%
  tnrs_match_names()

# Check the result
print(resolved_names)

# Filter only the matched species
matched_species <- resolved_names %>%
  filter(is.na(approximate_match) | approximate_match == FALSE)

# If you want to keep only successfully matched species (those with a valid ott_id)
matched_species_clean <- matched_species %>%
  filter(!is.na(ott_id))

# Check the cleaned matched species
print(matched_species_clean)

# First, rename the column correctly using rename() function
matched_species_clean <- matched_species_clean %>%
  rename(species_name = search_string) %>%  
  mutate(species_name = str_replace(species_name, "^(\\w)", toupper))

#Separate out taxa groups for plotting

#This data frame has been exported - read in and begin as below

#take relevant columns, same from the species data frame and left join
ott_ids <- read_csv("ott_id.csv") %>%
  mutate(
    species_name = str_replace(species_name, "^(\\w)", toupper),
    ott_id = as.numeric(ott_id))                                    
  
phylo <- left_join(ott_ids, species, by = "species_name")

#Algae
algae <- phylo %>%
  filter(taxon == "Macroalgae")

#Fish
fish <- phylo %>%
  filter(taxon == "Vertebrate")

#Inverts
inverts <- phylo %>%
  filter(taxon == "Invertebrate")



####### FISH ###########
#Let's improve this plot - Fishtree phylogeny is for rayed fish only: 
#Separate Actinopterygii out

# Split into two datasets
actinopterygii_fish <- fish %>% filter(class == "Actinopterygii")

# You can further split by slope category (declining vs. increasing)
# For Actinopterygii:
actin_declining <- actinopterygii_fish %>% filter(slope_category == "declining")
actin_increasing <- actinopterygii_fish %>% filter(slope_category == "increasing")

# Now retrieve the phylogenetic trees for each group
# Assume 'species_list' contains species names
phylo_actin <- fishtree_phylogeny(actinopterygii_fish$species_name)

#Only 303 of the 437 fish requested have been found

# Plot each phylogenetic tree with customization
# Function to enhance plotting of phylogenetic trees
plot_phylo_tree <- function(phylo_tree, title, species_data) {
  ggtree(phylo_tree, layout = "rectangular") +
    geom_tiplab(size = 2, align = TRUE, linetype = 'dashed') +  # Automatically uses tip labels
    geom_tippoint(aes(color = species_data$slope_category[match(phylo_tree$tip.label, species_data$species_name)]), size = 2) +  # Color by slope_category  
    theme_tree2() + 
    labs(title = title) +
    scale_color_manual(values = c("declining" = "red", "increasing" = "blue")) +
    theme(legend.position = "right", text = element_text(size = 12))  # Customize text and legend
}

# Plot Actinopterygii declining and increasing species separately
plot_actin_declining <- plot_phylo_tree(fishtree_phylogeny(actin_declining$species_name), "Actinopterygii (Declining)", actin_declining)
plot_actin_increasing <- plot_phylo_tree(fishtree_phylogeny(actin_increasing$species_name), "Actinopterygii (Increasing)", actin_increasing)

# Visualize the plots (optional) to check the output
plot_actin_declining
plot_actin_increasing

# Function to enhance plotting of phylogenetic trees
plot_phylo_tree <- function(phylo_tree, title, species_data) {
  # Check if phylo_tree is a valid tree object
  if (!inherits(phylo_tree, "phylo")) {
    stop("The provided phylo_tree is not a valid phylogenetic tree object.")
  }

  # Create the plot
  ggtree(phylo_tree, layout = "rectangular") +
    geom_tiplab(aes(label = .data$tip.label), size = 2, align = TRUE, linetype = 'dashed') +  # Use .data for better compatibility
    geom_tippoint(aes(color = species_data$slope_category), size = 2) +  # Color by slope_category
    theme_tree2() + 
    labs(title = title) +
    scale_color_manual(values = c("declining" = "red", "increasing" = "blue")) +
    theme(legend.position = "right", text = element_text(size = 12))  # Customize text and legend
}

# Plot Actinopterygii increasing species
plot_actin_increasing <- plot_phylo_tree(fishtree_phylogeny(actin_increasing$species_name), "Actinopterygii (Increasing)", actin_increasing)



################## Phylogenetic Least Squares analysis #######################
pgls_data <- phylo %>%
  filter(species_name %in% phylo) %>%
  left_join(covariates, by = site_code)  

# Convert your data into comparative data object (required for PGLS)
comp_data <- comparative.data(phy = phylo_tree, 
                              data = pgls_data, 
                              names.col = ("site_code"), 
                              vcv = TRUE, 
                              na.omit = FALSE)

# Check the comparative data object
print(comp_data)

# Perform PGLS analysis
# First attempt - slope as the response variable, body size and temperature as predictors
pgls_model <- pgls(slope ~ max_length + temperature, data = comp_data)

# Check the summary of the PGLS model
summary(pgls_model)

# If you want to visualize model diagnostics, you can plot residuals
par(mfrow = c(2, 2))
plot(pgls_model)

# Extract coefficients and p-values
coef(pgls_model)
summary(pgls_model)$coefficients

```
